---
title: "Practical Machine Learning : Prediction Assignment Writeup"
author: "Lakshya Sivaramakrishnan"
date: "Sunday, August 23, 2015"
output: html_document
---


**Summary**

As a part of the Practical Machine Learning course, we are conducting a prediction model experiment. The intent of this assignment is to predict the manner in which people exercise. Dataset : http://groupware.les.inf.puc-rio.br/har


**Background**

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal is  to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. 


**Data Cleaning**

Training and testing datasets were available. Data loading process was done as follows :

```{r}
pmlTrain<-read.csv("pml-training.csv", header=T, na.strings=c("NA", "#DIV/0!"))
pmlTest<-read.csv("pml-testing.csv", header=T, na.string=c("NA", "#DIV/0!"))
```


All the variables with at least one "NA" were excluded from the analysis. Variables related to time and user information were excluded for a total of 51 variables and 19622 class measurements. Same variables were maintained in the test data set (Validation dataset) to be used for predicting the 20 test cases provided.


```{r}
#NA exclusion
noNApmlTrain<-pmlTrain[, apply(pmlTrain, 2, function(x) !any(is.na(x)))] 
dim(noNApmlTrain)

#Time and user information variables exclusion
cleanpmlTrain<-noNApmlTrain[,-c(1:8)]
dim(cleanpmlTrain)

#Validation data set having clean information
cleanpmltest<-pmlTest[,names(cleanpmlTrain[,-52])]
dim(cleanpmltest)
```


**Data Partitioning and Prediction Process**

The cleaned downloaded data set was subset in order to generate a test set independent from the 20 cases provided set. Partitioning was performed to obtain a 75% training set and a 25% test set.

```{r}
library(caret)
inTrain<-createDataPartition(y=cleanpmlTrain$classe, p=0.75,list=F)
training<-cleanpmlTrain[inTrain,] 
test<-cleanpmlTrain[-inTrain,] 
#Training set dimensions
dim(training)

#Testing set dimensions
dim(test)
```


**Results and Conclusions**

Random forest trees were generated for the training dataset using cross-validation. Then, the generated algorithm was examined under the partitioned training set to examine the accuracy and estimated error of prediction. By using 51 predictors for five classes using cross-validation at a 5-fold, an accuracy of 99.2% with a 95% CI [0.989-0.994] was achieved accompanied by a Kappa value of 0.99.

```{r}
library(caret)
set.seed(13333)
fitControl2<-trainControl(method="cv", number=5, allowParallel=T, verbose=T)
rffit<-train(classe~.,data=training, method="rf", trControl=fitControl2, verbose=F)


predrf<-predict(rffit, newdata=test)
confusionMatrix(predrf, test$classe)

pred20<-predict(rffit, newdata=cleanpmltest)
pred20
```


**Submission**

On getting the prediction results for the 20 test cases, the below shown script was used to obtain single text files to be uploaded to the courses web site to comply with the submission assigment. 20 out of 20 hits also confirmed the accuracy of the obtained models.

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(pred20)
```